Future challenges for Research Software Engineering

### 2025-05-12 future_challenges_for_RSEng.tex
today's state

% Main part of your article
% =========================
\section{Introduction}

Research Software Engineering is currently an emerging field,
with an increasing integration into various aspects of research.
In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
We present here the results of a workshop discussion at the
deRSE25 conference in Karlsruhe\cite{Goth2025EndRSEng}, without
claiming to represent the entire deRSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in imagining the future.
In the following sections, we consider the evolution of the RSE definition,
the development of RSE education,
the impact of automation and generative artificial intelligence in RSE work,
the emerging ethical considerations,
and the transformations expected in the research organizations.
We then attempt to make long-term predictions
and we conclude the paper with our wishes and recommendations.

@@

\subsection{How have academic Institutions transformed by then?}
GC-NOTE: With longer-term funding and more trust to pick up longer term projects,
some RSEs can focus on code modernization, cleanup, and modularization,
which can pay off in faster and larger development in the future.
This is not possible by short-term funded solo RSEs.

\begin{whatis}{What do we want to have achieved}{academic}
We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects[FIXME: cite group paper].
We also hope, that specialized clusters form, with certain RSE units specializing in certain tasks like data management, HPC, or visualization.
At universities which have them, we want to see that researchers naturally accept them as part of their used services.
\end{whatis}

\subsubsection{Which challenges will still persist}
The workshop participants expect that there will still be budgeting and administrative hurdles to creating
permanent and well-funded RSE teams.
While we hope that the barrier to researchers diminishes, the workshop participants
still expect some work to do. We hope that the compartmentalization that occurs will help researchers reprioritize their workload.
An upcoming challenge will be that, with the emergence of more dedicated RSE units,
networking between them will come into the foreground, because there should be a strong need to share best practices
among institutions with successful RSE Teams.
On a more organisational level, we expect the sharing of models, to fit the different organisational constraints.

What are open questions here?
Organizations will have to come up with models for hiring people. While
a standardized RSE Master's makes it easier, it will be interesting to see if
RSE are still often hired by word of mouth or if the master's really makes it more comparable.

\begin{FramedParagraphWithFootnotes}
  GC-NOTE: New section on "New challenges":

  \begin{itemize}
    \item Networking of embedded RSEs and RSE units
    \item Everyone needs infrastructure: does everyone need to host their own infrastructure?
    \item Everyone hosts their infrastructure, but then everyone stays in silos? We need federated infrastructure.
  \end{itemize}
\end{FramedParagraphWithFootnotes}

@@

\subsection{What are new tasks of RTPs?}
 FG-NOTE: RSEs are one type of specialization of RTPs ins science. Do we see any other?
\begin{FramedParagraphWithFootnotes}
JN-NOTE:
New RSE specialisations might arise.
HPC-RSE and AI-RSE are quite recent developments,
whose appeal is stimulated by large public and private
investments in HPC facilities and AI technologies.
We could imagine new RSE specialisations to emerge in the fields
of quantum computing and federated online platforms.
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum
\end{FramedParagraphWithFootnotes}

\subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}

RSEs can either be embedded in research groups (belonging to the mid-level university staff),
or be members of larger RSE units. (TODO: Cite RSE groups paper)
The participants had the opinion that focusing on embedded RSEs would make onboarding new RSEs to existing projects easier.
This, however, comes with additional funding challenges,
as long-term funding is more difficult to acquire in lower levels of an organization.
Third-party funding can mitigate this issue in the mid-term.
However, the participants considered central RSE groups more likely in the future,
due to the fact that there are already examples and activities from various universities in this direction
(examples include the universities of Jena, Heidelberg, Göttingen).
We expect funding in central groups to evolve from initially temporary positions to permanent positions,
as the groups themselves prove themselves valuable.
Embedded RSEs and central groups will need to network with each other,
and potentially request and coordinate support from external RSEs.

@@

% Bibliography with BibTeX
% ========================
%\bibliographystyle{eceasst}
%\bibliography{bibliography/bibliography,extra}

>>>>

% Main part of your article
% =========================
\section{Introduction}
This opinion article stems from a Bof session at deRSE25 in Karlsruhe\cite{Goth2025EndRSEng}.
Therefore it stems from a discussion among the participants of this workshop and hence cannot claim that it represents the entire deRSE community.
Nevertheless we hope, that we succeeded in writing up something that
\begin{itemize}
\item is pleasant to read,
\item gives something to think about,
\item motivates engaging in work towards a better future.
\end{itemize}
In order to structure the discussion, we decided on an imaginable setting ten years from now.
In this world we will consider the dimensions of 
\begin{itemize}
\item Evolution of the term RSEs
\item The development of academic institutions
\item What will the impact of AI be to the work of RSEs?
\item How will academic education have transformed by then,
\item The Development of the RSE community,
\item Ethics and social considerations
\end{itemize}
In each aspect we will quickly summarise on what we want to have achieved by then, and what will be the challenges that arise from that.

@@

\subsection{How have academic Institutions transformed by then?}
Academia, while claiming to be innovative (cite WissZeitVG here), has organisationally often adopted time-scales for change that are more reminiscient to civil services like pension funds.
Nevertheless, the presence, even today, of initiatives like the HiddenRef hints at the potential for specialisation of research tasks. This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
We have alreadyseen some specialisation trends in recent years such as
the HPC-RSE and AI-RSE which are quite recent developments, whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
But we could imagine new RSE specialisations to emerge in the fields
of quantum computing, such as the Quantum-RSE, and in federated online platforms.
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum

On the other hand the existence of people with a certain specilised skillset opens up possiblities for pooling.
Universities will benefit from pooling these specialists into RSE units, and offer their services to researchers.
\begin{whatis}{What do we want to have achieved}{academic}
We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects[FIXME: cite group paper].
We also hope, that specialized clusters form that pool the skills of certain specialised RSEs, with certain RSE units specializing in certain tasks like data management, HPC, or visualization.
At universities which have them, we want to see that researchers naturally accept them as part of their used services.
\end{whatis}
Hopefully, these RSE-units are endowed with enough trust by the university boards that translates into stable funding.
With longer-term funding and more trust to pick up longer term projects,
some RSEs can focus on code modernization, cleanup, and modularization,
which can pay off in faster and larger development in the future.
This is an opportunity that would not possible by short-term funded solo RSEs.

\emph{Which challenges will still persist}\\
The workshop participants expect that there will still be budgeting and administrative hurdles to creating
permanent and well-funded RSE teams.
While we hope that the barrier to researchers diminishes, the workshop participants
still expect some work to do. We hope that the compartmentalization that occurs will help researchers reprioritize their workload.
An upcoming challenge will be that, with the emergence of more dedicated RSE units,
networking between them will come into the foreground, because there should be a strong need to share best practices
among institutions with successful RSE Teams.
On a more organisational level, we expect the sharing of models on how how to organise diiferent RSEs, to fit the different organisational constraints.
After that Organisations will have to come up with models for hiring people.
While a standardized RSE Master's makes it easier, it will be interesting to see if
RSE are still often hired by word of mouth within an organisation(``I have this PhD here, before their PostDoc they would be a good fit for the RSE department...'' ) or if the RSE master's degree ensures a hiring process on a more standardised level.

@@

\emph{New challenges}
If we follow the ideas laid out in (FIXME: cite RSE group paper), then in addition to university wide RSE units,
there will be embedded RSEs in the departments. Shaping and developing these network bonds will be a task for the future.
While networking embedded RSEs and RSE units is essential for building robust, sustainable research software infrastructure, but the question arises whether every group truly needs to host and manage its own infrastructure(FIXME: maybe cite this here: https://www.dfg.de/en/research-funding/funding-opportunities/programmes/infrastructure/lis/funding-opportunities/research-software-infrastructures).
When each team or institution hosts its own systems, this often leads to siloed environments where resources, expertise, and data are isolated, limiting collaboration and creating inefficiencies. Such silos can result in duplicated efforts, higher costs, and challenges in maintaining data integrity and governance, as each group independently manages similar infrastructure and workflows. While local control offers flexibility, it also increases complexity and makes it harder to share innovations or best practices across the wider research community. To overcome these challenges, federated infrastructure models are needed, enabling shared access, interoperability, and coordinated development while still allowing for some local autonomy(FIXME: maybe cite this here: https://eurise-network.github.io/).
Federated approaches break down silos, promote knowledge exchange, and ensure that infrastructure investments benefit the broader ecosystem rather than just isolated groups

\subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
RSEs can either be embedded in research groups (belonging to the mid-level university staff),
or be members of larger RSE units. (TODO: Cite RSE groups paper)
The participants had the opinion that focusing on embedded RSEs would make onboarding new RSEs to existing projects easier.
This, however, comes with additional funding challenges,
as long-term funding is more difficult to acquire in lower levels of an organization.
Third-party funding can mitigate this issue in the mid-term.
However, the participants considered central RSE groups more likely in the future,
due to the fact that there are already examples and activities from various universities in this direction
(examples include the universities of Jena, Heidelberg, Göttingen).
We expect funding in central groups to evolve from initially temporary positions to permanent positions,
as the groups themselves prove themselves valuable.
Embedded RSEs and central groups will need to network with each other,
and potentially request and coordinate support from external RSEs.

@@

% Bibliography with BibTeX
% ========================
\bibliographystyle{eceasst}
\bibliography{bibliography/bibliography,extra}
